---
title: "01_models"
output: html_document
---

Importing the libraries
```{r}
library(mikropml)
library(dplyr)
library(ggplot2)
library(tidyr)
```

Loading the preprocessed dataset and filtering
```{r}
input_file <- file.path(data_dir, "prepared_data_final.rds")
dat <- readRDS(input_file)

# outcome first
dat$dx <- factor(dat$dx)

# keep only the OTU columns (numeric features)
otu_cols <- grep("^\\d|^OTU|^err|^srr", names(dat), value = TRUE)

dat <- dat %>%
  select(dx, all_of(otu_cols))

# force numeric safety
dat <- dat %>%
  mutate(across(-dx, as.numeric))

# remove NA rows
dat <- na.omit(dat)

table(dat$dx)
```

...
```{r}
model_preprocess <- c("nzv", "corr", "center", "scale")
```

Checking the data structure
```{r}
cat("Outcome distribution:\n")
print(table(dat$dx))

cat("Missing values:\n")
print(sum(is.na(dat)))
```

Linear model
```{r}
results_glmnet <- run_ml(
  dataset = dat,
  method = "glmnet",
  outcome_colname = "dx",
  seed = seed,
  preProcess = model_preprocess,
  cv_times = 10,
  tune_length = 20
)

results_glmnet$performance
```
```

Random Forest
```{r}
results_rf <- run_ml(
  dataset = dat,
  method = "rf",
  outcome_colname = "dx",
  seed = seed,
  preProcess = model_preprocess,
  cv_times = 50 # Reduced for the tutorial run
)
```

Checking the resutls
```{r}
cat("\n--- Training Results Summary ---\n")
cat("GLMNET Performance (Internal CV AUC):", results_glmnet$performance$cv_metric_AUC, "\n")
cat("GLMNET Performance (Test Set AUC):", results_glmnet$performance$AUC, "\n")
cat("\nRF Performance (Internal CV AUC):", results_rf$performance$cv_metric_AUC, "\n")
cat("RF Performance (Test Set AUC):", results_rf$performance$AUC, "\n")
```

Combining the results...
```{r}
model_list <- list(
  glmnet = results_glmnet,
  rf = results_rf
)

fix_perf_types <- function(df) {
  df %>%
    mutate(across(where(is.character) & !c(method), as.numeric))
}

perf_glmnet <- fix_perf_types(model_list$glmnet$performance)
perf_rf <- fix_perf_types(model_list$rf$performance)

combined_perf <- bind_rows(perf_glmnet, perf_rf)

colnames(combined_perf) <- tolower(colnames(combined_perf))
```

Selecting the needed results for comparison
```{r}
print(
  combined_perf %>%
  select(method, cv_metric_auc, accuracy, mean_f1)
)
```

Combining the results for comparison
```{r}
perf_long <- combined_perf %>%
  select(method, cv_metric_auc, accuracy, mean_f1) %>% #prAUC,
  pivot_longer(-method, names_to = "metric", values_to = "value")

p_comp <- ggplot(perf_long, aes(x = method, y = value, fill = method)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~metric, scales = "free_y") +
  theme_bw() +
  labs(title = "Model Performance Comparison", 
       subtitle = "Higher values indicate better performance",
       y = "Score", x = "ML Method") +
  scale_fill_brewer(palette = "Set1")

p_comp
```

Creating helper functions
```{r}
# Helper function to get sens/spec for plotting
get_curves_data <- function(result, method_name) {
  calc_model_sensspec(result$trained_model, result$test_data, "dx") %>%
    mutate(method = method_name)
}

roc_prc_glmnet <- get_curves_data(model_list$glmnet, "glmnet")
roc_prc_rf <- get_curves_data(model_list$rf, "rf")
combined_curves <- bind_rows(roc_prc_glmnet, roc_prc_rf)
```

Creating the ROC curve
```{r}
p_roc <- combined_curves %>%
  ggplot(aes(x = specificity, y = sensitivity, color = method)) +
  geom_line(linewidth = 1) +
  geom_abline(intercept = 1, slope = 1, linetype = "dashed", color = "grey") +
  scale_x_reverse() +
  theme_bw() +
  coord_fixed() +
  labs(title = "Receiver Operating Characteristic (ROC) Curve",
       x = "Specificity (False Positive Rate)",
       y = "Sensitivity (True Positive Rate)")

p_roc
```

Creating the PRC curve
```{r}
p_prc <- combined_curves %>%
  rename(recall = sensitivity) %>%
  ggplot(aes(x = recall, y = precision, color = method)) +
  geom_line(linewidth = 1) +
  theme_bw() +
  coord_fixed() +
  labs(title = "Precision-Recall Curve (PRC)",
       x = "Recall (Sensitivity)",
       y = "Precision")

p_prc
```

