---
title: "00_preparation"
output: html_document
---

#' 01_Data_Preparation.R
#' 
#' This script performs initial data cleaning for microbial metagenomics 
#' machine learning analysis. It covers the first phase of the lab practical:
#' Data Preparation.
#'
#' Steps:
#' 1. Cleaning Names and Columns
#' 2. Standardizing Feature Data Types
#' 3. Saving a prepared dataset for leakage-safe model training
```{r}
library(dplyr)
library(janitor)
```

```{r Load configuration and dependencies}
source(file.path(base_dir, "config.R"))

otu <- read.delim(
  file.path(data_dir, "otu_table.xls"),
  sep = "\t",
  header = TRUE,
  check.names = FALSE
)
metadata <- read.delim(
  file.path(data_dir, "sample_data.xls"),
  sep = "\t",
  header = TRUE,
  check.names = FALSE
)
```


```{r Ensure output directories exist before logging or saving artifacts}
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(logs_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figures_dir, recursive = TRUE, showWarnings = FALSE)
```


```{r Setup Logging and Seed}
library(tibble)

otu <- otu %>%
  column_to_rownames("featureid")

otu <- as.data.frame(t(otu))
otu$sample_id <- rownames(otu)

```

```{r}
head(otu)
```


# 1. Cleaning Names and Columns
# Novice Tip: Special characters or spaces in column names can break R functions.
# We use janitor::clean_names() to ensure all headers are snake_case and ASCII-safe.
```{r}
cleaned_data <- metadata %>%
  inner_join(otu, by = "feature_id")
```

```{r}
cleaned_data$dx <- as.factor(cleaned_data$dx)
```


# 2. Standardizing Feature Data Types
# Novice Tip: Models require numeric inputs. We ensure all OTU columns are numeric.
# In this example dataset they already are, but we apply this as a best practice.
# We leave dx as character because mikropml handles outcome conversion internally.
```{r}
cleaned_data <- cleaned_data %>%
  janitor::clean_names()

cleaned_data <- cleaned_data %>%
  mutate(across(-SampleID, as.numeric))
```


# Summary of changes
```{r}
cat("\n--- Preprocessing Summary ---\n")
cat("Sample count:", nrow(cleaned_data), "\n")
cat("Feature count:", ncol(cleaned_data) - 1, "\n")
cat("Outcome counts:\n")
print(table(cleaned_data$dx))
```


# Save intermediate output
```{r}
output_file <- file.path(results_dir, "prepared_data.rds")
saveRDS(cleaned_data, file = output_file)

cat("\nSuccess: Prepared data saved to results/prepared_data.rds\n")
cat("Data Preparation Phase Complete.\n")

sink()
```
