---
title: "01_prediction"
output: html_document
---
#' 02_Model_Training.R
#' 
#' This script performs machine learning model training and experimental design.
#' It covers the second phase of the lab practical:
#' Model Training and Experimental Design.
#'
#' Steps:
#' 1. Data Partitioning (Internal vs. External Validation)
#' 2. Leakage-safe preprocessing inside model training
#' 3. k-Fold Cross-Validation (Hyperparameter Tuning)
#' 4. Model Execution (Baseline: Logistic Regression)
#' 5. Model Execution (Non-linear: Random Forest)
#' 6. Parallelization for Speed

```{r}
library(mikropml)
library(dplyr)
library(ggplot2)
library(tidyr)
```


# Load configuration and dependencies
```{r}
source(file.path(base_dir, "config.R"))
```

```{r}
# 0. Setup Logging, Seed, and Parallelization
log_file <- file.path(logs_dir, "02_model_training.log")
sink(log_file, append = FALSE, split = TRUE)
set.seed(seed)

cat("Starting Model Training Phase...\n")

# Use multiple cores for training speed.
# Novice Tip: Training models can be computationally expensive. 
# Parallelization allows R to train multiple versions of the model simultaneously.
n_cores <- 4 # Adjust based on environment; students may use more.
future::plan(future::multicore, workers = n_cores)
cat("Parallel processing enabled with", n_cores, "cores.\n")
```

```{r}
input_file <- file.path(results_dir, "prepared_data_final.rds")
dat <- readRDS(input_file)

# outcome first
dat$dx <- factor(dat$dx)

# keep ONLY OTU columns (numeric features)
otu_cols <- grep("^\\d|^OTU|^err|^srr", names(dat), value = TRUE)

dat <- dat %>%
  select(dx, all_of(otu_cols))

# force numeric safety
dat <- dat %>%
  mutate(across(-dx, as.numeric))

# remove NA rows safely
dat <- na.omit(dat)

table(dat$dx)
```



```{r}
# 2. Leakage-safe learned preprocessing
# caret estimates these operations from training data/resamples only, then applies
# the fitted transformation during prediction.
model_preprocess <- c("nzv", "corr", "center", "scale")
cat("Learned preprocessing inside caret::train():", 
    paste(model_preprocess, collapse = ", "), "\n")
```
```{r}
cat("Outcome distribution:\n")
print(table(dat$dx))

cat("Missing values:\n")
print(sum(is.na(dat)))

cat("Non-numeric columns (should only be dx):\n")
print(sapply(dat, class))
```
```{r}
cat("\nStructure check:\n")
str(dat)

cat("\nAny non-numeric predictors?\n")
print(sapply(dat[-1], class) |> unique())
```


```{r}
# 3. Model Training Strategy: Logistic Regression (glmnet)
# Novice Tip: Logistic regression with L2 regularization (Ridge) is a great baseline.
# It is fast, interpretable, and less prone to overfitting than complex models.
cat("\nStep 3: Training Logistic Regression (glmnet)...\n")
cat("Strategy: 80/20 Train-Test split, 5-fold Cross-Validation (repeated 2x for this tutorial)\n")

# run_ml() handles partitioning and CV automatically.
results_glmnet <- run_ml(
  dataset = dat,
  method = "glmnet",
  outcome_colname = "dx",
  seed = seed,
  preProcess = model_preprocess,
  cv_times = 2 # Reduced for the tutorial run; should be 100 in real research
)
```

```{r}
# 4. Model Training Strategy: Random Forest (rf)
# Novice Tip: Random Forest is a non-linear ensemble method. It's robust 
# but can take much longer to train.
cat("\nStep 4: Training Random Forest (rf)...\n")
results_rf <- run_ml(
  dataset = dat,
  method = "rf",
  outcome_colname = "dx",
  seed = seed,
  preProcess = model_preprocess,
  cv_times = 2 # Reduced for the tutorial run
)
```

```{r}
# 4. Inspecting Initial Results
cat("\n--- Training Results Summary ---\n")
cat("GLMNET Performance (Internal CV AUC):", results_glmnet$performance$cv_metric_AUC, "\n")
cat("GLMNET Performance (Test Set AUC):", results_glmnet$performance$AUC, "\n")
cat("\nRF Performance (Internal CV AUC):", results_rf$performance$cv_metric_AUC, "\n")
cat("RF Performance (Test Set AUC):", results_rf$performance$AUC, "\n")

# Novice Tip: If CV AUC is 1.0 but Test AUC is 0.5, the model is OVERFITTING.
# It has memorized the training data but cannot generalize to new samples.
```


```{r}
# 5. Save Model Results
output_file <- file.path(results_dir, "model_results.rds")
model_list <- list(
  glmnet = results_glmnet,
  rf = results_rf
)
saveRDS(model_list, file = output_file)

cat("\nSuccess: Trained models saved to results/model_results.rds\n")
cat("Model Training Phase Complete.\n")
```

#' 03_Model_Evaluation.R
#' 
#' This script evaluates the trained machine learning models and generates 
#' comparative figures and tables.
#' It covers the third phase of the lab practical:
#' Model Evaluation and Performance Metrics.
#'
#' Steps:
#' 1. Aggregating Performance Metrics
#' 2. Visualizing Performance Comparison (AUC, prAUC)
#' 3. Plotting ROC and Precision-Recall Curves
#' 4. Exporting Summary Tables

```{r}
# 0. Setup Logging and Seed
log_file <- file.path(logs_dir, "03_model_evaluation.log")
sink(log_file, append = FALSE, split = TRUE)
set.seed(seed)

cat("Starting Model Evaluation Phase...\n")
```

```{r}
# 1. Load Trained Models
input_file <- file.path(results_dir, "model_results.rds")
if (!file.exists(input_file)) {
  stop("Error: model_results.rds not found. Run 02_model_training.R first.")
}
model_list <- readRDS(input_file)
```

```{r}
# 2. Combine Performance Data
# We extract the 'performance' dataframe from each model result.
# Novice Tip: Sometimes R packages return "NA" as a string instead of a missing value.
# We ensure all metrics are numeric before combining.
fix_perf_types <- function(df) {
  df %>%
    mutate(across(where(is.character) & !c(method), as.numeric))
}

perf_glmnet <- fix_perf_types(model_list$glmnet$performance)
perf_rf <- fix_perf_types(model_list$rf$performance)

combined_perf <- bind_rows(perf_glmnet, perf_rf)

colnames(combined_perf) <- tolower(colnames(combined_perf))
```

```{r}
names(combined_perf)
str(combined_perf)
```


```{r}
cat("\nStep 1: Comparative Performance Table\n")

f1_col <- if ("mean_f1" %in% names(combined_perf)) "mean_f1" else "f1"

print(
  combined_perf %>%
    dplyr::select(method, cv_metric_auc, accuracy, !!sym(f1_col))
)

write.csv(combined_perf,
          file.path(results_dir, "performance_comparison.csv"),
          row.names = FALSE)
```

```{r}
# 3. Visualizing Comparison
cat("\nStep 2: Generating performance comparison plots...\n")

perf_long <- combined_perf %>%
  select(method, cv_metric_auc, accuracy, !!sym(f1_col)) %>% #prAUC,
  pivot_longer(-method, names_to = "metric", values_to = "value")

p_comp <- ggplot(perf_long, aes(x = method, y = value, fill = method)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~metric, scales = "free_y") +
  theme_bw() +
  labs(title = "Model Performance Comparison", 
       subtitle = "Higher values indicate better performance",
       y = "Score", x = "ML Method") +
  scale_fill_brewer(palette = "Set1")

ggsave(file.path(figures_dir, "model_comparison.png"), p_comp, width = 8, height = 6)
```


```{r}
# 4. ROC and Precision-Recall Curves
cat("\nStep 3: Calculating ROC and PRC curves...\n")

# Helper function to get sens/spec for plotting
get_curves_data <- function(result, method_name) {
  calc_model_sensspec(result$trained_model, result$test_data, "dx") %>%
    mutate(method = method_name)
}

roc_prc_glmnet <- get_curves_data(model_list$glmnet, "glmnet")
roc_prc_rf <- get_curves_data(model_list$rf, "rf")
combined_curves <- bind_rows(roc_prc_glmnet, roc_prc_rf)
```

```{r}
# Plot ROC Curve
# Novice Tip: ROC curve shows the trade-off between Sensitivity and Specificity.
# A curve closer to the top-left corner is better.
p_roc <- combined_curves %>%
  ggplot(aes(x = specificity, y = sensitivity, color = method)) +
  geom_line(linewidth = 1) +
  geom_abline(intercept = 1, slope = 1, linetype = "dashed", color = "grey") +
  scale_x_reverse() +
  theme_bw() +
  coord_fixed() +
  labs(title = "Receiver Operating Characteristic (ROC) Curve",
       x = "Specificity (False Positive Rate)",
       y = "Sensitivity (True Positive Rate)")

ggsave(file.path(figures_dir, "roc_curve.png"), p_roc, width = 6, height = 6)
```


```{r}
# Plot Precision-Recall Curve
# Novice Tip: PRC is often better for imbalanced biological data.
p_prc <- combined_curves %>%
  rename(recall = sensitivity) %>%
  ggplot(aes(x = recall, y = precision, color = method)) +
  geom_line(linewidth = 1) +
  theme_bw() +
  coord_fixed() +
  labs(title = "Precision-Recall Curve (PRC)",
       x = "Recall (Sensitivity)",
       y = "Precision")

ggsave(file.path(figures_dir, "prc_curve.png"), p_prc, width = 6, height = 6)
```

