---
title: "olivia"
output: html_document
---

Loading the necessary packages:

```{r load-packages}
library(phyloseq)
library(tidyverse)
library(ape)
library(tibble)
library(readxl)
library("ggplot2"); packageVersion("ggplot2")
```

Loading the data:
```{r load-input-files}
species_dir <- "./data/D003093_Colitis,_Ulcerative/species/"
genus_dir <- "./data/D003093_Colitis,_Ulcerative/genus/"

otu_df <- read.delim(
  file.path(species_dir, "otu_table.xls"),
  check.names = FALSE
)

tax_df <- read.delim(
  file.path(species_dir, "tax_table.xls"),
  check.names = FALSE
)

sample_df <- read.delim(
  file.path(species_dir, "sample_data.xls"),
  check.names = FALSE
)


head(otu_df)
```

Creating tree:
```{r}
tree <- ape::read.tree(file.path(species_dir, "tree.nwk"))
```


Preparing phyloseq input:
```{r prepare-phyloseq-inputs}
otu_mat <- otu_df %>%
  column_to_rownames("featureid") %>%
  as.matrix()

tax_mat <- tax_df %>%
  column_to_rownames("#featureid") %>%
  as.matrix()

sample_tab <- sample_df %>%
  column_to_rownames("#SampleID")
```

Creating the phyloseq object:
```{r}
ps <- phyloseq(
  otu_table(otu_mat, taxa_are_rows = TRUE),
  tax_table(tax_mat),
  sample_data(sample_tab),
  phy_tree(tree)
)

ps
```

Making it a long format:
```{r}
ps_long <- psmelt(ps)
head(ps_long)
```

```{r}
disease_ID <- subset_samples(
  ps,
  Disease.MESH.ID =="D003093")
```

Taxanomic rank:
```{r}
ps_genus <- tax_glom(ps,
                     taxrank = "Genus",
                     NArm = TRUE)

taxa_names(ps_genus) <- make.unique(
  as.character(tax_table(ps_genus)[, "Genus"])
)

#tax_table(ps_genus)
```

```{r}
ps_genus <- tax_glom(ps,
                     taxrank = "Species",
                     NArm = TRUE)

taxa_names(ps_genus) <- make.unique(
  as.character(tax_table(ps_genus)[, "Species"])
)

#tax_table(ps_genus)
```

```{r}
plot_bar(ps, fill = "Phylum")
```

Plot heatmap for subsets of age:
```{r}
ps_female <- subset_samples(ps, Sex == "Female")
ps_male <- subset_samples(ps, Sex == "Male")

plot_heatmap(ps_female)

plot_heatmap(ps_male)
```

Plot heatmaps for subsets of recent antibiotics use:
```{r}
ps_ab_y <- subset_samples(ps, Recent.antibiotics.use == "Y")
ps_ab_n <- subset_samples(ps, Recent.antibiotics.use == "N")

plot_heatmap(ps_ab_n)

plot_heatmap(ps_ab_y)
```

<https://joey711.github.io/phyloseq/index.html>

Filtering low-abundance taxa or low-depth samples:
```{r}

```


Agglomerating at different taxonomic ranks and comparing plots:
```{r}

```

Shannon richness:
```{r}
shannon <- estimate_richness(
  ps,
  measures = "Shannon"
)

head(shannon)
```

Comparing richness plots:
```{r}
plot_richness(
  ps,
  measures = "Shannon"
)

plot_richness(
  ps,
  measures = "Shannon",
  x = "Country"
)

plot_richness(
  ps,
  measures = "Shannon",
  x = "Sex"
)

plot_richness(
  ps,
  measures = "Shannon",
  x = "Host.age"
)

plot_richness(
  ps,
  measures = "Shannon",
  x = "Recent.antibiotics.use"
)

```

```{r}
plot_richness(
  ps,
  measures = "Shannon",
  x = "Country",
  color = "Recent.antibiotics.use"
  )

plot_richness(
  ps,
  measures = "Shannon",
  x = "Host.age",
  color = "Recent.antibiotics.use"
  )

plot_richness(
  ps,
  measures = "Shannon",
  x = "Sex",
  color = "Recent.antibiotics.use"
  )
```

Observed richness:
```{r}
ps_pa <- transform_sample_counts(ps, function(x) {ifelse(x > 0, 1, 0)})

observed <- estimate_richness(
  ps_pa,
  measures = "Observed"
)

head(observed)
```

Building an alpha-diversity table:
```{r}
shannon <- estimate_richness(
  ps,
  measures = "Shannon"
) %>%
  rownames_to_column("Sample")

# Calculate observed richness
observed <- estimate_richness(
  ps_pa,
  measures = "Observed"
) %>%
  rownames_to_column("Sample")

# Extract sample metadata
sample_df <- sample_data(ps) %>%
  data.frame() %>%
  rownames_to_column("Sample")

# Combine diversity estimates and metadata
alpha_diversity <- inner_join(
  shannon,
  observed,
  by = "Sample"
) %>%
  inner_join(
    sample_df,
    by = "Sample"
  )

head(alpha_diversity)
```

Kruskal-Shannon test:
```{r}
# Country, Sex, Recent.antibiotics.use, Host.age
kruskal_shannon <- kruskal.test(
  Shannon ~ Recent.antibiotics.use,
  data = alpha_diversity
)

kruskal_shannon
```

Shannon box-plot:
```{r}
ggplot(alpha_diversity, aes(x = Recent.antibiotics.use, y = Shannon)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  theme_bw()

ggplot(alpha_diversity, aes(x = Sex, y = Shannon)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  theme_bw()

ggplot(alpha_diversity, aes(x = Country, y = Shannon)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  theme_bw()
```

Total abundance per sample:
```{r}
sample_depth <- sample_sums(ps)

alpha_diversity$reads <- sample_depth[
  alpha_diversity$Sample
]

ggplot(alpha_diversity, aes(x = reads, y = Shannon)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
```

Correlations test:
```{r}
cor.test(
  alpha_diversity$reads,
  alpha_diversity$Shannon,
  method = "spearman"
)
```

